Kernel k-Groups and Hartigan's Method

### 2018-06-23 energy_clustering/pami18/energy_clustering.tex
maria comments

\IEEEtitleabstractindextext{%
\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by 
Newton's gravitational potential in classical mechanics, and it provides
a model-free 
hypothesis test for equality of distributions. In its original form, energy
statistics was formulated in 
Euclidean spaces. More recently, it 
was generalized to metric spaces of negative type. 
In this paper, we consider a formulation for the clustering problem
using a weighted version of 
energy statistics in spaces of negative type.
We show that this approach leads to a
quadratically constrained
quadratic program in the associated kernel space, establishing
connections with graph partitioning problems and
kernel methods in unsupervised machine learning.
To find local solutions of such an optimization problem, 
we propose an extension of Hartigan's method to kernel spaces.
Our method has the same computational cost
as kernel k-means algorithm, which is based on Lloyd's heuristic, but 
our numerical results show an improved performance,
specially in high dimensions.
\end{abstract}

@@

Clustering is an important unsupervised learning problem and 
has such a long history in statistics and machine learning, making it
impossible to mention all important contributions in a short space. 
Perhaps, the most used method is k-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of iteratively computing
the means of each cluster and then assigning points to
the cluster with closest center. The only statistical 
information about each cluster comes from its mean, making the method sensitive 
to outliers. Nevertheless, k-means works very well when data is 
linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
another very common approach, providing more flexibility than k-means, 
however, it still makes strong assumptions about the distribution of 
the data.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen. A well-known kernel based clustering method
is kernel k-means, which is precisely k-means 
formulated in the feature space \cite{Girolami}. 
Furthermore, kernel k-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic.
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

The predominant approach to clustering %, in Euclidean or kernel spaces,
is based on Lloyd's method.
However, there is an old heuristic
due to Hartigan \cite{Hartigan1975,Hartigan1979}, which apparently did
not receive the deserved attention, 
and goes as follows: for each data point, simply assign it to a cluster
in an optimal way such that a loss function is minimized.
While Lloyd's method only iterates if some cluster contains a point
that is closer to the mean of another cluster, Hartigan's method may iterate
even if that is not the case, and moreover, it takes into account the motion
of the means resulting from the reassignments. In this sense, Hartigan's
method may potentially escape local minima of Lloyd's method.
In the Euclidean case, this was shown to be the case \cite{Telgarsky}.
Moreover, the advantages of Hartigan's over Lloyd's method was verified
empirically \cite{Telgarsky,Slonin}, however, 
although it was observed to be as fast as Lloyd's method, no complexity
analysis was provided.

@@

Another contribution of this paper is to extend 
Hartigan's method to  kernel spaces. 
To the best of our knowledge, such an extension was not previously considered.
Since this approach was motivated by energy statistics and  
\cite{Kgroups} considered the Euclidean case, we call the 
proposed method \emph{kernel k-groups}.
We show that kernel k-groups 
has the same complexity as kernel k-means algorithm, however,
our numerical results provide compelling evidence that 
kernel k-groups 
is more accurate and robust, specially in high dimensions.

Using the standard kernel defined by energy
statistics, our experiments illustrate 
that kernel k-groups
is able to perform accurately on data coming from 
very different distributions, contrary to k-means and GMM, for instance.
More specifically, our method performs 
closely to k-means and GMM on normally distributed data, however,
it is significantly better on data that 
is not normally distributed. 
Its superiority in high dimensions
is striking, being more accurate than k-means and GMM 
even on Gaussian settings. We also illustrate the advantages of
kernel k-groups on a real dataset.

>>>>

\IEEEtitleabstractindextext{%
\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by 
Newton's gravitational potential in classical mechanics, and it provides
a model-free 
hypothesis test for equality of distributions. In its original form, energy
statistics was formulated in 
Euclidean spaces. More recently, it 
was generalized to metric spaces of negative type. 
In this paper, we consider a formulation for the clustering problem
using a weighted version of 
energy statistics in spaces of negative type.
We show that this approach leads to a
quadratically constrained
quadratic program in the associated kernel space, establishing
connections with graph partitioning problems and
kernel methods in unsupervised machine learning.
To find local solutions of such an optimization problem, 
we propose an extension of Hartigan's method to kernel spaces.
Our method has the same computational cost
as kernel k-means algorithm, which is based on Lloyd's heuristic, but 
our numerical results show an improved performance,
especially in high dimensions.
\end{abstract}

@@

Clustering is an important unsupervised learning problem and 
has a long history in statistics and machine learning, making it
impossible to mention all important contributions in a short space. 
Perhaps, the most used method is k-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of iteratively computing
the means of each cluster and then assigning points to
the cluster with closest center. The only statistical 
information about each cluster comes from its mean, making the method sensitive 
to outliers. Nevertheless, k-means works very well when data is 
linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
another very common approach, providing more flexibility than k-means; 
however, it still makes strong assumptions about the distribution of 
the data.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A Mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen. A well-known kernel based clustering method
is kernel k-means, which is precisely k-means 
formulated in the feature space \cite{Girolami}. 
Furthermore, kernel k-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic.
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Besides Lloyd's approach to clustering %, in Euclidean or kernel spaces,
there is an old heuristic
due to Hartigan \cite{Hartigan1975,Hartigan1979} that
goes as follows: for each data point, simply assign it to a cluster
in an optimal way such that a loss function is minimized.
While Lloyd's method only iterates if some cluster contains a point
that is closer to the mean of another cluster, Hartigan's method may iterate
even if that is not the case, and moreover, it takes into account the motion
of the means resulting from the reassignments. In this sense, Hartigan's
method may potentially escape local minima of Lloyd's method.
In the Euclidean case, this was shown to be the case \cite{Telgarsky}.
Moreover, the advantages of Hartigan's over Lloyd's method was verified
empirically \cite{Telgarsky,Slonin}. However, 
although it was observed to be as fast as Lloyd's method, no complexity
analysis was provided.

@@

Another contribution of this paper is to extend 
Hartigan's method to  kernel spaces. 
To the best of our knowledge, such an extension was not previously considered.
Since this approach was motivated by energy statistics and  
\cite{Kgroups} considered the Euclidean case, we call the 
proposed method \emph{kernel k-groups}.
We show that kernel k-groups 
has the same complexity as kernel k-means algorithm, however,
our numerical results provide compelling evidence that 
kernel k-groups 
is more accurate and robust, especially in high dimensions.

Using the standard kernel defined by energy
statistics, our experiments illustrate 
that kernel k-groups
is able to perform accurately on data coming from 
very different distributions, contrary to k-means and GMM, for instance.
More specifically, our method performs 
closely to k-means and GMM on normally distributed data, while
it is significantly better on data that 
is not normally distributed. 
Its superiority in high dimensions
is striking, being more accurate than k-means and GMM 
even in Gaussian settings. We also illustrate the advantages of
kernel k-groups on real data.
